---
title: "Exploratory analysis of Global Terrorism Database"
author: "Andrew Ba Tran"
date: "10/31/2018"
output:
  html_document:
    code_folding: hide
---

This analysis is based on the [Global Terrorism Database](https://www.start.umd.edu/gtd/) paired with the ideology categorizations.

This supplementary data is from  Miller, Erin, 2017, "Global Terrorism Database Ideological Motivations of Terrorism in the United States", and originally hosted at [Harvard Dataverse](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/SACQNK). 

It is a supplemental file that adds categories to all the incidents tracked by the [GTD](https://www.start.umd.edu/gtd/), such as Environmental extremism, Left-wing extremism, Right-wing extremism, Religious extremism, etc.

The categories only are as recent as 2016. The Post attempted to mimic the ideological categorization from START.

Eventually we didn't feel comfortable mixing these data sets so we went back and categorized every incident in the United States [ourselves](http://wpinvestigative.github.io/global_terrorism_database_analysis/rmd/post_analysis.html).

Visit the [repo for this project](http://www.github.com/wpinvestigative/global_terrorism_database_analysis) for more information/data.


```{r setup, warning=F, message=F}
# replace NAs with blanks when displaying data in the markdown table
options(knitr.kable.NA = '')

library(tidyverse)
library(readxl)
library(knitr)

# We cannot rehost GTD's raw data files as per their Terms of Agreement
# If you're looking to reproduce our work, there are some extra steps
# It requires downloading their data and placing it into the `raw_data` folder in this project folder

# Load Raw Data -- instructions
# Download the raw data files from GTD 

raw1 <- read_excel("../raw_data/gtd_70to95_0718dist.xlsx", sheet=1)
raw2 <- read_excel("../raw_data/gtd_96to13_0718dist.xlsx", sheet=1)
raw3 <- read_excel("../raw_data/gtd_14to17_0718dist.xlsx", sheet=1)
raw4 <- read_excel("../raw_data/gtd1993_0718dist.xlsx", sheet=1)
rawData <- rbind(raw1, raw2, raw3, raw4)

# Loading ideology categorizations from START for incidents between 1970 and 2016
db <- load("../raw_data/globalterrorismdb_USideology_1970-2016.RData")

# Loading the Post's ideology categorizations for incidents in 2017
db2017 <- read_csv("../raw_data/ideology_2017.csv")

# Joining the two categorization data sets together
x <- rbind(x, db2017)
x$eventid <- as.numeric(x$eventid)

# Joining the GTD incident data with the supplemental ideology categorization data
md_h <- inner_join(rawData, x)
```


## Fatalities in the U.S. over time

```{r fatalities_by_year, warning=F, message=F, fig.width=8, fig.height=4}
md_h %>% group_by(iyear) %>% 
  summarize(fatalities=sum(nkill, na.rm=T)) %>% 
  select(year=iyear, fatalities) %>% 
  ggplot() + 
  aes(x=year, y=fatalities) +
  geom_bar(stat="identity")
```


## Fatalities in the U.S. over time (excluding 2001 figures)

```{r fatalities_by_year2, warning=F, message=F, fig.width=8, fig.height=4}
md_h %>% group_by(iyear) %>% 
  summarize(fatalities=sum(nkill, na.rm=T)) %>% 
  filter(iyear!=2001) %>% select(year=iyear, fatalities) %>% 
  ggplot() + 
  aes(x=year, y=fatalities) +
  geom_bar(stat="identity")

```


## Total incidents in the U.S. over time 

```{r incidents_by_year, warning=F, message=F, fig.width=8, fig.height=4}
md_h %>% group_by(iyear) %>% 
  summarize(incidents=n())  %>% 
  select(year=iyear, incidents) %>% 
  ggplot() + 
  aes(x=year, y=incidents) +
  geom_bar(stat="identity")

```

```{r category_prep, warning=F, message=F}
# Recreating new data frames based on categorizations
# There is probably a more efficient way to do this but ¯\_(ツ)_/¯ this works

env <- filter(md_h, ENVIRONMENTAL==1) %>% 
  mutate(category="ENVIRONMENTAL")

lw <- filter(md_h, LEFT.WING==1) %>% 
  mutate(category="LEFT WING")

rw <- filter(md_h, RIGHT.WING==1) %>% 
  mutate(category="RIGHT WING")

rws <- filter(md_h, RIGHT.WING.SOVEREIGN==1) %>% 
  mutate(category="RIGHT WING SOVEREIGN")

rwa <- filter(md_h, RIGHT.WING.ANTIGOV==1) %>% 
  mutate(category="RIGHT WING ANTIGOV")

r <- filter(md_h, RELIGIOUS==1) %>% 
  mutate(category="RELIGIOUS")

rc <- filter(md_h, RELIGIOUS.CHRISTIAN==1) %>% 
  mutate(category="RELIGIOUS CHRISTIAN")

rj <- filter(md_h, RELIGIOUS.JEWISH==1) %>% 
  mutate(category="RELIGIOUS JEWISH")

ri <- filter(md_h, RELIGIOUS.ISLAMIC==1) %>% 
  mutate(category="RELIGIOUS ISLAMIC")

is <- filter(md_h, ISLAMIC.SHIA==1) %>% 
  mutate(category="ISLAMIC SHIA")

isu <- filter(md_h, ISLAMIC.SUNNI==1) %>% 
  mutate(category="ISLAMIC SUNNI")

ns <- filter(md_h, NATIONALIST.SEPARATIST==1) %>% 
  mutate(category="NATIONAL SEPARATIST")

si <- filter(md_h, SINGLE==1) %>% 
  mutate(category="SINGLE")

newdf <- rbind(env, lw, rw, rws, rwa, r, rc, rj, ri, is, isu, ns, si)
```

## Incidents since 1972

```{r incidents_total_chart_1970, fig.width=9, fig.height=6, warning=F, message=F, echo=F}
newdf %>% group_by(category, iyear) %>% 
  summarize(fatalities=sum(nkill, na.rm=T), incidents=n()) %>% 
  filter(iyear>1972) %>% 
  ggplot() + 
  aes(x=iyear, y=incidents) +
  geom_bar(stat="identity") +
  facet_wrap(~category, ncol=3)
```


## Incidents since 2002

```{r incidents_total_chart, fig.width=9, fig.height=6, warning=F, message=F}
newdf %>% group_by(category, iyear) %>% 
  summarize(fatalities=sum(nkill, na.rm=T), incidents=n()) %>% 
  filter(iyear>2001) %>% 
  ggplot() + 
  aes(x=iyear, y=incidents) +
  geom_bar(stat="identity") +
  facet_wrap(~category, ncol=3)
```


```{r incidents_total_table, fig.width=9, fig.height=6, warning=F, message=F}
newdf %>% group_by(category, iyear) %>% 
  summarize(incidents=n()) %>% 
  filter(iyear>2001) %>% 
  spread(iyear, incidents) %>% 
  kable()
```


## Fatalities since 2002

```{r fatalities_total_chart, fig.width=9, fig.height=6, warning=F, message=F}
newdf %>% group_by(category, iyear) %>% 
  summarize(fatalities=sum(nkill, na.rm=T), incidents=n()) %>% 
  filter(iyear>2001) %>% 
  ggplot() + 
  aes(x=iyear, y=fatalities) +
  geom_bar(stat="identity") +
  facet_wrap(~category, ncol=3)
```


```{r fatalities_total_table, fig.width=9, fig.height=6, warning=F, message=F}

newdf %>% group_by(category, iyear) %>% 
  summarize(fatalities=sum(nkill, na.rm=T)) %>% 
  filter(iyear>2001) %>% 
  spread(iyear, fatalities) %>% 
  kable()
```


## Incidents since 2002 by group

```{r incidents_total_chart_group, fig.width=9, fig.height=9, warning=F, message=F}

newdf %>% group_by(gname, iyear) %>% 
  summarize(fatalities=sum(nkill, na.rm=T), incidents=n()) %>% 
  filter(iyear>2001) %>% 
  ggplot() + 
  aes(x=iyear, y=incidents) +
  geom_bar(stat="identity") +
  facet_wrap(~gname, ncol=4)
```


```{r incidents_total_table_group, fig.width=9, fig.height=9, warning=F, message=F}
newdf %>% group_by(gname, iyear) %>% 
  summarize(incidents=n()) %>% 
  filter(iyear>2001) %>% 
  spread(iyear, incidents) %>% 
  kable()
```


## Fatalities since 2002 by group

```{r fatalities_total_chart_group, fig.width=9, fig.height=9, warning=F, message=F}
newdf %>% group_by(gname, iyear) %>% 
  summarize(fatalities=sum(nkill, na.rm=T), incidents=n()) %>% 
  filter(iyear>2001) %>% 
  ggplot() + 
  aes(x=iyear, y=fatalities) +
  geom_bar(stat="identity") +
  facet_wrap(~gname, ncol=4)
```


```{r fatalities_total_table_group, fig.width=9, fig.height=6, warning=F, message=F}
newdf %>% group_by(gname, iyear) %>% 
  summarize(fatalities=sum(nkill, na.rm=T)) %>% 
  filter(iyear>2001) %>% 
  spread(iyear, fatalities) %>% 
  kable()
```


